---
title: "Euler's method"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

Differential equations describe how one term changes with respect to another term. One ecology example is how a population changes over time.

$$
\frac{dN}{dt} = rN
$$

Where $dN$ is a small change in population size $N$ with a small change in time $t$. $r$ is the growth rate (births - deaths).

We can extend this model to include carrying capacity (now a logistic model).

$$
\frac{dN}{dt} = rN (1- \frac{N}{K})
$$

Euler's method is a technique to solve differential equations. 
Using that method, we can solve the equation to get this:

$$
N_{t+dt} = N_{t} + [rN_{t} (1-\frac{N_{t}}{K}) ]dt
$$

We are going to try and simulate the logistic model using Euler’s method.

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
```


```{r}
get_N <- function(K, N0, r, t) {
  (K * N0) / (N0 + (K - N0) * exp(- r * t))
}
```


Let's say we know that at the initial time point $N$ is 10,
and that $r$ is 0.1 and K is 30.

```{r}
tibble(
    t = seq(1, 50, 1)
  ) %>% 
  mutate(
    N = get_N(K = 30, N0 = 10, r = 0.1, t = t)
    ) %>% 
  ggplot(aes(x = t, y = N)) +
  geom_point() +
  geom_hline(yintercept = 30, linetype = 2)
```

