---
title: "Matrix community models"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r}
library("tidyverse")
library("patchwork")
```

> Matrix community models are sets of single-species matrix population models that are linked together to capture community-wide dynamics. [Lytle & Tonkin, 2023](https://doi.org/10.1038/s44185-023-00031-5)

You can link the species matrices by including a density assumption -
density dependence is a result of abundances of other species.

```{r}
t_mat <- matrix(     
  c(
    0, 0, 10,
    0.2, 0.5, 0,
    0, 0.2, 0.8
  ),
  nrow = 3, ncol = 3, byrow = T,
  dimnames = (list(c("seed", "juvenile", "adult"),
                  NULL))
)

t_mat
```

|  | seed | juvenile | adult |
|:---------|:-----|:------|:------|
| **seed**     | seed remaining seed | from juvenile to seed | from adult to seed |
| **juvenile** | from seed to juvenile | juvenile remaining juvenile |  from adult to juvenile   |
| **adult**    | from seed to adult  | from juvenile to adult |   adult remaining adult    |



Starting with 30 juvenile plants

```{r}
abund_init <- c(0, 30, 0, 0)

abund_init
```



```{r}
yr_1 <- t_mat %*% abund_init 

yr_1
```

```{r}
n_years <- 30

all_years <- matrix(0, 
                    nrow = nrow(t_mat), 
                    ncol = n_years + 1)

all_years[,1] <- yr_1

for(t in 2:(n_years + 1)){   
  all_years[,t] <-  t_mat %*% all_years[,t-1]
}

all_years
```

```{r}
all_years %>% 
  as_tibble(name_repair = "unique") %>% 
  rowid_to_column(var = "stage") %>% 
  pivot_longer(cols = contains("V"),
               names_to = "time") %>% 
  mutate(stage = recode(stage, 
                        "seeds",
                        "juveniles",
                        "adults"),
         time = as.numeric(str_remove(time, pattern = "V"))) %>% 
  ggplot(aes(x = time, y = value, 
             colour = stage, group = stage)) +
  geom_line()
```
