---
title: "Equalization and Stabilization"
author: "Notes from Mike"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here are some notes on how to determine equalization and stabilization from both unstructured and structured models. *Equalization* is the idea that species have equivalent intrinsic rates of increase. *Stabilization* is the idea that density-dependent processes operate to return to a population to an equlibirium, prevent exponential increase/decreaase and potentially allow species to coexist.

We can use the *net reproductive rate* (net change in population size from generation/time step to the next) as way to vizualize these ideas of equalization and stablization. Here's it all for the dynamics of a single species ($N$) described by logistic growth/feedbacks:

\[
\frac{dN}{dt}=rN\left[1-\frac{N}{K}\right]
\]

where $r$ is the intrinsic rate of increase and $K$ is the carrying capacity. The net growth rate for simple unstructured populations is $\frac{1}{N} \frac{dN}{dt}$ so for the logistic model:

\[
\frac{1}{N}\frac{dN}{dt}=r-\frac{rN}{K}
\]

This is a straight-line - intercept $r$ - which represents the equalization processes - and slope $\frac{r}{K}$ which represents the strength of density-dependence - the stablization processes.

For a structured life-history - say with seedlings ($S$) and adults trees ($A$) - with the density-dependence acting on seedling mortality - we can write a set of coupled equations:

\[
\frac{dS}{dt}=rA(t)-log(1+aS(t))S(t)-mS(t)
\]

\[
\frac{dA}{dt}=mS(t)-\mu A(t)
\]

where $r$ is the seedling production rate, $log(1+aS(t))$ is a function describing density dependent seedling mortality, $m$ is seedling maturation rate and $\mu$ is adult tree mortality rate.

To determine net growth rate from a structured life-history, we work out the Jacobian matrix (contributions of all of life stage to changes in the dynamics of each life stage - these are partial derivatives) - say the contribution of adults to changes in seedlings is $\frac{\partial{\left[dS/dt\right]}}{\partial{A}}$.

The Jacobian for the seedling-tree model is:

\[
\left(\begin{array}{cc}
-\lambda-m-\frac{a}{1+aS} & r \\
m & -\lambda-\mu
\end{array}\right)
\]

where $\lambda$ are eigenvalues associated with the square matrix and, for us, the dominant eigenvalue represents the net growth rate associated with the structured life history.

By taking the determinant and solving this matrix for $\lambda$ we get:

\[
\lambda= \frac{-m-\frac{a}{1+aS}-\mu}{2}\pm\sqrt{\left(m+\frac{a}{1+aS}+\mu\right)^2-4\left(\frac{a\mu}{1+aS}+m(\mu-r)\right)}
\]

We can graph the positive valuves of lambda (net growth rate) to illustrate ideas of equalization and stabilization for two different specces


```{r}
#set model for species 1

m=0.1 # maturation rate
a=0.1 # strength of density dependent seedling mortality
r=5.0 # seedling recruitment rate
mu=0.001 # adult tree mortality rate

S1=seq(0,100,length=1000)

trm1=-m-(a/(1+a*S1))-mu
trm2= trm1^2
trm3=m*(mu-r)*((a*mu)/(1+a*S1))

lambda1=(trm1/2)+sqrt(trm2-4*trm3) 

#----------------------------------

#set model for species 1

m=0.1 # maturation rate
a=0.05 # strength of density dependent seedling mortality
r=5.0 # seedling recruitment rate
mu=0.001 # adult tree mortality rate

S2=seq(0,100,length=1000)

trm1a=-m-(a/(1+a*S2))-mu
trm2a= trm1^2
trm3a=m*(mu-r)*((a*mu)/(1+a*S2))
         
lambda2=(trm1a/2)+sqrt(trm2a-4*trm3a) 

plot(seq(0,100,length=10),seq(0.05,0.12,length=10),type="n",xlab="Density of Seedlings",ylab="Net Growth Rate (lambda)",bty="l")

lines(S1,lambda1,col='blue',lwd=2)
lines(S2,lambda2,col='red',lwd=2)
         
```

Equalization - for the red species (species 2), it has the highest expected net growth rate (intercept on the y-axis) and might be expected to win out in competition.

Stabilization - both species experience similar density dependence strengths and this could promote coexistence.
