---
title: "Structured models"
author: "Notes from Mike"
date: "2025-05-01"
output: html_document
---


This R Markdown document explains how a structured modelling framework could/can be designed to investigate the role of shared enemies on tree coexistence (and Jansen/Connell maintenance of diversity arguments).

Following the results that tree growth follows Gompertz-type growth with stage/age mortality, we can define a set of linked differential equations to mimic delayed growth dynamics:

\[
\frac{dx_0}{dt}=\lambda-\mu x_0 - x_0(t)
\]

\[
\frac{dx_i}{dt}=x_{i-1}(t) - \mu x_{i-1}(t) - x_i(t) \quad \text{i=1,...,n}
\]

\[
\frac{dF}{dt}=x_n(t) -\mu F(t) 
\]

The first equation describes the growth dynamics with a constant input (recruitment) ($\lambda$) and (exponentially-distributed) mortality ($\mu$) this growth dynamics is then coupled through a set of linked equations (representing stages/ages) to model the changes in growth to a final stage $\frac{dF}{dt}$.

To illustrate the dynamics, the following set of code solves a system where there 2 steps (2 linked equations) to describe the Gompertz growth dynamics:

```{r}
rm(list=ls())

#set the number of iterations (obs), integration step (h),  and start time (v)

obs=1000
v=0.0
h=0.05

#set the number of equations and model structure

ns=2
nvar=ns+1

#define the output array

x1=array(dim=c(obs,nvar))

y=array(dim=(nvar))

derivs=function(x,y){
  
  lambda=2.0
  mu=0.05
  
  yx[1]=lambda-mu*y[1] - y[1]
  
  for(i in 1:ns){
  yx[i+1]=y[i]-mu*y[i]-y[i+1]}

  yx[nvar]=y[ns]-mu*y[nvar]
  
  yx
}


#set the initial conditions 

for(i in 1:nvar)
y[i]=0.0

y[1]=0.1
y[nvar]=0.1



#arrays for the RK4 integrator (see below)

yt=array(1:nvar)
dyt=array(1:nvar)
dym=array(1:nvar)
yout=array(1:nvar)
yx=array(1:nvar)
eqn=array(1:nvar)


#4th order runge-kutta function (the workhorse for numerical integration)

rk4=function(dydx,y,t,n){
  for(jj in	1:n)
    yt[jj]=y[jj]+dydx[jj]*h/2
  dyt=derivs(t+h/2,yt)
  for(jj in 1:n)	
    yt[jj]=y[jj]+h/2*dyt[jj]
  dym=derivs(t+h/2,yt)
  for(jj in 1:n)
    yt[jj]=y[jj]+h*dym[jj]
  dym=dyt+dym
  dyt=derivs(t+h,yt)
  for(jj in 1:n)
    yout[jj]=y[jj]+(h/6)*(dydx[jj]+dyt[jj]+2*dym[jj])
  yout
}

#define output arrays

t1=array(1:obs)
x1=array(dim=c(obs,nvar))

#iterate the model 


for(i in 1:obs){
  eqn=derivs(v,y)
  y=rk4(eqn,y,v,nvar)
  v=v+h
  x1[i,]=y
  t1[i]=v
  
}

#plot some results

plot(seq(0,max(t1),length=10),seq(0,max(x1[,nvar]+5),length=10),type="n",bty="l",xlab="Time",ylab="Growth")
lines(t1,x1[,nvar],lwd=2)

```

The model is now iterated for three different lengths of steps describing the Gompertz growth (2,4 & 8). (Code is hidden).


```{r,echo=FALSE}
rm(list=ls())

#set the number of iterations (obs), integration step (h),  and start time (v)

obs=1000
v=0.0
h=0.05

#set the number of equations and model structure
res=array(0,dim=c(obs,3))
stps=c(2,4,8)

for(kjk in 1:3){
ns=stps[kjk]
nvar=ns+1

#define the output arrays

x1=array(dim=c(obs,nvar))

y=array(dim=(nvar))

derivs=function(x,y){
  
  lambda=2.0
  mu=0.05
  
  yx[1]=lambda-mu*y[1] - y[1]
  
  for(i in 1:ns){
  yx[i+1]=y[i]-mu*y[i]-y[i+1]}

  yx[nvar]=y[ns]-mu*y[nvar]
  
  yx
}


#set the initial conditions 

for(i in 1:nvar)
y[i]=0.0

y[1]=0.1
y[nvar]=0.1



#arrays for the RK4 integrator (see below)

yt=array(1:nvar)
dyt=array(1:nvar)
dym=array(1:nvar)
yout=array(1:nvar)
yx=array(1:nvar)
eqn=array(1:nvar)


#4th order runge-kutta function (the workhorse for numerical integration)

rk4=function(dydx,y,t,n){
  for(jj in	1:n)
    yt[jj]=y[jj]+dydx[jj]*h/2
  dyt=derivs(t+h/2,yt)
  for(jj in 1:n)	
    yt[jj]=y[jj]+h/2*dyt[jj]
  dym=derivs(t+h/2,yt)
  for(jj in 1:n)
    yt[jj]=y[jj]+h*dym[jj]
  dym=dyt+dym
  dyt=derivs(t+h,yt)
  for(jj in 1:n)
    yout[jj]=y[jj]+(h/6)*(dydx[jj]+dyt[jj]+2*dym[jj])
  yout
}

#define output arrays

t1=array(1:obs)
x1=array(dim=c(obs,nvar))

#iterate the model 


for(i in 1:obs){
  eqn=derivs(v,y)
  y=rk4(eqn,y,v,nvar)
  v=v+h
  x1[i,]=y
  t1[i]=v
  
}
res[,kjk]=x1[,nvar]
}

```

Plotting the results for the different lengths of steps describing the Gompertz growth (2 (black),4 (red) & 8 (orange)) for a set of fixed parameters ($\lambda=2.0$, $\mu=0.5$) captures different types of growth dynamics: 

```{r, echo=FALSE}
tt=seq(1,50,length=length(res[,1]))
#plot some results

plot(seq(0,max(tt),length=10),seq(0,max(res[,]+1),length=10),type="n",bty="l",xlab="Time",ylab="Growth")
lines(tt,res[,1],lwd=2,col="black")
lines(tt,res[,2],lwd=2,col="red")
lines(tt,res[,3],lwd=2,col="orange")

```
